

\section*{\bibtitleauthorEn}
% Publications on the dissertation contents
%Author's publications on the dissertation topic


\begin{enumerate}
	
	\item \textbf{Shirokikh B.}, Zakazov I., Chernyavskiy A., Fedulova I., Belyaev M. First U-Net layers contain more domain specific information than the last ones // Lecture Notes in Computer Science (including subseries Lecture Notes in Artificial Intelligence and Lecture Notes in Bioinformatics). -- 2020. -- Vol. 12444. -- p. 117--126. -- DOI: 10.1007/978-3-030-60548-3\_12 (Indexed in Scopus);
	
    \item Saparov T., Kurmukov A., \textbf{Shirokikh B.}, Belyaev M. Zero-Shot Domain Adaptation in CT Segmentation by Filtered Back Projection Augmentation // Lecture Notes in Computer Science (including subseries Lecture Notes in Artificial Intelligence and Lecture Notes in Bioinformatics). -- 2021. -- Vol. 13003. -- p. 243--250. -- DOI: 10.1007/978-3-030-88210-5\_24 (Indexed in Scopus);
    
    \item Goncharov M.\footnotemark[1], Pisov M.\footnotemark[1], Shevtsov A.\footnotemark[1], \textbf{Shirokikh B.}\footnotemark[1], Kurmukov A., Blokhin I., Chernina V., Solovev A., Gombolevskiy V., Morozov S., Belyaev M. CT-Based COVID-19 triage: Deep multitask learning improves joint identification and severity quantification // Medical Image Analysis. -- 2021. -- Vol. 71. -- p. 102054. -- DOI: 10.1016/j.media.2021.102054 \footnotetext[1]{shared contribution} (Indexed in Scopus, Q1 journal, IF: 10.7);
    
    \item  Zakazov I.\footnotemark[2], \textbf{Shirokikh B.}\footnotemark[2], Chernyavskiy A., Belyaev M. Anatomy of domain shift impact on U-Net layers in MRI segmentation // Lecture Notes in Computer Science (including subseries Lecture Notes in Artificial Intelligence and Lecture Notes in Bioinformatics). -- 2021. -- Vol. 12903. -- p. 211--220. -- DOI: 10.1007/978-3-030-87199-4\_20 \footnotetext[2]{shared contribution} (Indexed in Scopus, proceedings of A conference);
    
    \item Shimovolos S., Shushko A., Belyaev M., \textbf{Shirokikh B.} Adaptation to CT Reconstruction Kernels by Enforcing Cross-Domain Feature Maps Consistency // Journal of Imaging. -- 2022. -- Vol. 8, Is. 9. -- p. 234. -- DOI: 10.3390/jimaging8090234 (Indexed in Scopus, Q2 journal, IF: 2.7);
    
    \item \textbf{Shirokikh B.}, Dalechina A., Shevtsov A., Krivov E., Kostjuchenko V., Durgaryan A., Galkin M., Golanov A., Belyaev M. Systematic clinical evaluation of a deep learning method for medical image segmentation: radiosurgery application // IEEE Journal of Biomedical and Health Informatics. -- 2022. -- Vol. 26, Is. 7. -- p. 3037--3046. -- DOI: 10.1109/JBHI.2022.3153394 (Indexed in Scopus, Q1 journal, IF: 6.7);
    
    \item Vasiliuk A., Frolova D., Belyaev M., \textbf{Shirokikh B.} Limitations of out-of-distribution detection in 3d medical image segmentation // Journal of Imaging. -- 2023. -- Vol. 9, Is. 9. -- p. 191. -- DOI: 10.3390/jimaging9090191 (Indexed in Scopus, Q2 journal, IF: 2.7);
    
    \item Certificate of state registration of computer programs No. 2021612647(RU), Belyaev M., Pisov M., \textbf{Shirokikh B.}, 20 Feb 2021 (in Russian).
        
\end{enumerate}

    


\printbibliography[heading=subbibliography]